---
title: "STAT545 Assignment 2"
author: "Alex"
date: "September 24, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initializing Gapminder Dataset and Libraries
```{r, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(gapminder)
```

## "Smell test" the data
```{r}
typeof(gapminder)
```

The gapminder object is a list. 

```{r}
class(gapminder)

```

It's class is a tibble, which is a type of data frame. 

```{r}
dim(gapminder)
```

Using the`dims()` function we get that there are 1706 rows (observations) for 6 columns (variables). We could also get size information using the `ncol()` or `nrow()` functions

```{r}
summary(gapminder)
class(gapminder$country)
class(gapminder$continent)
class(gapminder$year)
class(gapminder$lifeExp)
class(gapminder$pop)
class(gapminder$gdpPercap)
```

Using `summary()` we can get a quick overview of the variables in *gapminder* and surmise the data type of each variable, but we can also make sure by calling the `class()` function.

From this we get the following data types in *gapminder*:

| Variable | Data Type |
|---------|-------|
| country | factor |
| continent | factor |
| year | integer |
| lifeExp | numeric |
| pop | integer |
| gdpPercap | numeric |





